---
title: "ST2195 Coursework Project - Part 2"
author: "Malek"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Packages

```{r packages}
library(tidyverse)
library(readr)
library(ggplot2)
library(knitr)
library(broom)
```

# Paths and files

```{r paths}
path <- "C:/Users/touns/OneDrive/Desktop/Bsc Data Science and Business Analytics/ST2159 Programming for Data Science/Coursework prog/Datasets"
files <- c("3.csv", "4.csv", "5.csv", "6.csv", "7.csv")
```

# Part 2(a): Best times and days to minimise delays

## Data cleaning and preprocessing

```{r q2a-clean}
cols_a <- c("Year", "Month", "DayofMonth", "DayOfWeek", "CRSDepTime", "ArrDelay", "DepDelay", "Cancelled")

hour_delay_list <- list()
day_delay_list  <- list()

for (file in files) {
  df <- read_csv(file.path(path, file), col_select = all_of(cols_a), show_col_types = FALSE)
  
  df <- df %>%
    filter(Cancelled == 0) %>%
    drop_na(ArrDelay, DepDelay, CRSDepTime, DayOfWeek) %>%
    mutate(Hour = CRSDepTime %/% 100) %>%
    filter(Hour >= 0, Hour <= 23)
  
  hour_delay <- df %>%
    group_by(Year, Hour) %>%
    summarise(ArrDelay = mean(ArrDelay), .groups = "drop")
  
  day_delay <- df %>%
    group_by(Year, DayOfWeek) %>%
    summarise(ArrDelay = mean(ArrDelay), .groups = "drop")
  
  hour_delay_list[[length(hour_delay_list) + 1]] <- hour_delay
  day_delay_list[[length(day_delay_list) + 1]] <- day_delay
}

hour_delay <- bind_rows(hour_delay_list) %>% arrange(Year, Hour)
day_delay  <- bind_rows(day_delay_list) %>% arrange(Year, DayOfWeek)
```

## Best hour and best day by year

```{r q2a-best}
best_hour <- hour_delay %>%
  group_by(Year) %>%
  slice_min(order_by = ArrDelay, n = 1, with_ties = FALSE) %>%
  ungroup()

best_day <- day_delay %>%
  group_by(Year) %>%
  slice_min(order_by = ArrDelay, n = 1, with_ties = FALSE) %>%
  ungroup()

kable(best_hour, caption = "Best scheduled departure hour by year")
kable(best_day, caption = "Best day of week by year")
```

## Figure 1: Average arrival delay by scheduled departure hour

```{r fig1-hour, fig.width=8, fig.height=5}
ggplot(hour_delay, aes(x = Hour, y = ArrDelay, color = factor(Year))) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Average Arrival Delay by Scheduled Departure Hour (2003-2007)",
    x = "Scheduled Departure Hour",
    y = "Average Arrival Delay (minutes)",
    color = "Year"
  ) +
  theme_minimal()
```

## Figure 2: Average arrival delay by day of week

```{r fig2-day, fig.width=8, fig.height=5}
ggplot(day_delay, aes(x = DayOfWeek, y = ArrDelay, color = factor(Year))) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:7) +
  labs(
    title = "Average Arrival Delay by Day of the Week (2003-2007)",
    x = "Day of Week",
    y = "Average Arrival Delay (minutes)",
    color = "Year"
  ) +
  theme_minimal()
```

## Appendix A.1: Hour-by-year detailed plot

```{r appendix-a1, fig.width=8, fig.height=7}
ggplot(hour_delay, aes(x = Hour, y = ArrDelay)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Average Arrival Delay by Scheduled Departure Hour for Each Year (2003-2007)",
    x = "Scheduled Departure Hour",
    y = "Average Arrival Delay (minutes)"
  ) +
  theme_minimal()
```

## Appendix A.2: Day-by-year detailed plot

```{r appendix-a2, fig.width=8, fig.height=7}
ggplot(day_delay, aes(x = DayOfWeek, y = ArrDelay)) +
  geom_line(color = "darkred") +
  geom_point(color = "darkred") +
  facet_wrap(~ Year, scales = "free_y") +
  scale_x_continuous(breaks = 1:7) +
  labs(
    title = "Average Arrival Delay by Day of the Week for Each Year (2003-2007)",
    x = "Day of Week",
    y = "Average Arrival Delay (minutes)"
  ) +
  theme_minimal()
```

# Part 2(b): Aircraft age analysis

## Data cleaning and preprocessing

```{r q2b-clean}
cols_b <- c("Year", "TailNum", "ArrDelay", "DepDelay", "Cancelled")

flights_list <- list()

for (file in files) {
  df <- read_csv(file.path(path, file), col_select = all_of(cols_b), show_col_types = FALSE)
  
  df <- df %>%
    filter(Cancelled == 0) %>%
    drop_na(ArrDelay, DepDelay) %>%
    mutate(TotalDelay = ArrDelay + DepDelay)
  
  flights_list[[length(flights_list) + 1]] <- df
}

flights <- bind_rows(flights_list)

planes <- read_csv(file.path(path, "plane-data.csv"), show_col_types = FALSE) %>%
  rename(TailNum = tailnum) %>%
  select(TailNum, year) %>%
  mutate(year = as.numeric(year)) %>%
  distinct(TailNum, .keep_all = TRUE)

flights_planes <- flights %>%
  left_join(planes, by = "TailNum") %>%
  mutate(PlaneAge = Year - year) %>%
  filter(!is.na(PlaneAge), PlaneAge >= 0, PlaneAge <= 40)

age_delay <- flights_planes %>%
  group_by(PlaneAge) %>%
  summarise(ArrDelay = mean(ArrDelay), .groups = "drop")
```

## Figure 3: Average arrival delay as a function of aircraft age

```{r fig3-age, fig.width=8, fig.height=5}
ggplot(age_delay, aes(x = PlaneAge, y = ArrDelay)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Average Arrival Delay as a Function of Aircraft Age (2003-2007)",
    x = "Aircraft Age (years)",
    y = "Average Arrival Delay (minutes)"
  ) +
  theme_minimal()
```

## Appendix A.3: Distribution of arrival delays

```{r appendix-a3, fig.width=8, fig.height=5}
ggplot(flights_planes, aes(x = ArrDelay)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black") +
  coord_cartesian(xlim = c(-50, 200)) +
  labs(
    title = "Distribution of Arrival Delays (2003-2007)",
    x = "Arrival Delay (minutes)",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Appendix A.4: Average arrival delay by aircraft age

```{r appendix-a4, fig.width=8, fig.height=5}
ggplot(age_delay, aes(x = PlaneAge, y = ArrDelay)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Arrival Delay by Aircraft Age (2003-2007)",
    x = "Aircraft Age (years)",
    y = "Average Arrival Delay (minutes)"
  ) +
  theme_minimal()
```

# Part 2(c): Logistic regression for flight diversion

## Data cleaning and model fitting

```{r q2c-model}
cols_c <- c(
  "Year", "Month", "DayOfWeek",
  "CRSDepTime", "CRSArrTime",
  "Distance", "DepDelay",
  "UniqueCarrier", "Diverted", "Cancelled"
)

results_c <- list()
coef_by_year <- list()
carrier_frames <- list()

for (file in files) {
  df <- read_csv(file.path(path, file), col_select = all_of(cols_c), show_col_types = FALSE)
  
  df <- df %>%
    filter(Cancelled == 0) %>%
    drop_na(Month, DayOfWeek, CRSDepTime, CRSArrTime, Distance, DepDelay, UniqueCarrier, Diverted) %>%
    mutate(
      DepHour = CRSDepTime %/% 100,
      ArrHour = CRSArrTime %/% 100
    ) %>%
    filter(DepHour <= 23, ArrHour <= 23)
  
  year_value <- df$Year[1]
  
  diverted_df <- df %>% filter(Diverted == 1)
  non_diverted_df <- df %>% filter(Diverted == 0)
  
  if (nrow(diverted_df) == 0) next
  
  n_sample <- min(nrow(non_diverted_df), max(20000, 20 * nrow(diverted_df)))
  non_div_sample <- non_diverted_df %>% slice_sample(n = n_sample)
  
  df_model <- bind_rows(diverted_df, non_div_sample) %>% slice_sample(prop = 1)
  
  model <- glm(
    Diverted ~ Month + DayOfWeek + DepHour + ArrHour + Distance + DepDelay,
    data = df_model,
    family = binomial(link = "logit")
  )
  
  coef_tbl <- tidy(model) %>%
    filter(term != "(Intercept)") %>%
    arrange(estimate)
  
  coef_tbl$Year <- year_value
  results_c[[length(results_c) + 1]] <- coef_tbl
  coef_by_year[[as.character(year_value)]] <- coef_tbl
  carrier_frames[[length(carrier_frames) + 1]] <- df %>% select(Year, UniqueCarrier, Diverted)
}

coef_df_report <- bind_rows(results_c)
carrier_df <- bind_rows(carrier_frames)
```

## Figure 4: Logistic regression coefficients for predicting flight diversions

```{r fig4-logit, fig.width=8, fig.height=5}
example_year <- sort(unique(coef_df_report$Year))[1]

coef_df_report %>%
  filter(Year == example_year) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = paste0("Logistic Regression Coefficients for Predicting Flight Diversions (", example_year, ")"),
    x = "Feature",
    y = "Coefficient"
  ) +
  theme_minimal()
```

## Figure 5: Airlines with the highest average flight diversion rates

```{r fig5-carriers, fig.width=8, fig.height=5}
top_carriers <- carrier_df %>%
  group_by(UniqueCarrier) %>%
  summarise(DiversionRate = mean(Diverted), .groups = "drop") %>%
  arrange(desc(DiversionRate)) %>%
  slice_head(n = 10)

ggplot(top_carriers, aes(x = reorder(UniqueCarrier, DiversionRate), y = DiversionRate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Airlines with the Highest Average Flight Diversion Rates (2003-2007)",
    x = "Carrier",
    y = "Average Diversion Rate"
  ) +
  theme_minimal()
```

## Appendix A.5: Logistic regression coefficients by year

```{r appendix-a5, fig.width=10, fig.height=8}
ggplot(coef_df_report, aes(x = reorder(term, estimate), y = estimate)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ Year, scales = "free_y") +
  labs(
    title = "Logistic Regression Coefficients by Year (2003-2007)",
    x = "Feature",
    y = "Coefficient"
  ) +
  theme_minimal()
```
